Shapes analysis for time series

### 2024-10-25 NEURIPS24/paper.tex
add github

% citep
% \section{TODO}
% %De ce que je comprends , les gens font surtout des variational auto encoder, ou du masking
% %Mentionner que la méthode proposé peut être utilisé avec des réseaux
% %\sam{Approche NTK, fédéré personalisation, projet avec Axel, Lien entre continuous normalizing flow}
% Plan intro: 
% \begin{itemize}
%     \item Representation of patterns in time series has several applications in biology, but halso for classification, deep learning finds features automaticly.
%      Taking or not taking into account time can be done with DTW. However, either the representation is the result of a black box and need post analysis to be understand,
%       which is difficult to apply statiscal tools on it, or the representation is desgned by hand to select relevant features, but the capacity of analysis is limited by the knowledge of the designer.
%      \item In shape analysis, representation of shapes for statistical analysis in a long time issue, which has been solved mainly by the LDDMM framework. Representing shapes as diffeomorpshims of a common shape. 
%      \item However, this tool has not been yet applied on time series, even if there is already some tools developped for oriented curves. 
%      \item In this work, we propose to apply LDDMM to time series and to show its interest to derive an interpretable and unsupervised representation of patterns in time series.
%      We proposed some tools developped in the shape analysis community to the researcher interested in machine learning and time series ones.
%      \item First, we introduce our assumptions regarding the dataset and the main problem of diffeomorpshim learning. Then, we show how to solve it with LDDMM .
%       We expose how to apply it to time series by selecting wisely a RKHS kernel, motivating it with a representation theorem.
%       experiments on synthetic data are presented to show that the parameter of representation are identifiable when the time seriess are not too sharp and the kernels parameter well tuned. 
%       Then, we sh-ow how to apply the method on an unsupervised case of moouse respeitaory dataset and a supervised one on ... .

% \end{itemize}

% One option to tackle this issue is to derive ...
% feature representation of time series which depends on the problem at hand.
% which is parti... -> which is of prime interest in ...
\vspace{-1ex}
\section{Introduction}
\vspace{-1ex}
%https://arxiv.org/abs/1811.00075
%Appliqué sur les "UWaveGestureLibrary","ArticularyWordRecognition","Cricket" pour l'archive UEA
%"ArrowHead", "BME", "ECG200","FacesUCR","GunPoint" dans le cas univarié

%Idée de parler d'inter individualité, spécialiser davantage
%k shape, shapelets , ye2009time

@@

Our work explicitly focuses on learning shape-based representation of time series.
First, we propose to view the shape of a time series not merely as its curve $\{s_t:\eqsp t\in\msi\}$, but as its graph $\msg(s)=\{(t,s(t)):\eqsp t\in \msi\}$.
   Then, building on the shape analysis literature \cite{beg2005computing,vaillant2004statistics}, we adopt the Large Deformation Diffeomorphic Metric Mapping (LDDMM) framework \cite{beg2005computing,vaillant2004statistics} to analyze these graphs.
    The core idea is to represent each element $\msg(s^j)$ of a dataset $(s^j)_{j\in[N]}$ as the transformation of a reference graph $\msg(\mathbf{s}_0)$ by a diffeomorphism $\phi_j$, i.e. $\msg(s^j) \sim \phi_j . \msg(\mathbf{s}_0)$ .
    The diffeomorphism $\phi_j$ is learned by integrating an ordinary differential equation parameterized by a Reproducing Kernel Hilbert Space (RKHS).
     The parameters $(\alpha_j)_{j\in[N]}$ encoding the diffemorphisms $(\phi_j)_{j\in[N]}$ yield the representation features of the graphs $(\msg(s^j))_{j\in[N]}$.
     Finally, these shape-encoding features can be used as inputs to any statistical or machine-learning model.

However, a graph time series transformation by a general diffeomorphism is not always a graph time series, see e.g. \Cref{fig:diffeo}, thus a graph time series is more than a simple curve \cite{glaunes2008large}.
 Our contributions arise from this observation: we specify the class of diffeomorphisms to consider and show how to learn them.
  This change is fruitful in representing transformations of time series graphs as illustrated in \Cref{fig:transport}.
   %In this regard, this work is not an application of LDDMM to 1D curves \cite{glaunes2008large}.

     %In particular, the method can handle \textit{irregularly sampled} time series with \textit{variable sizes}.

    %  Our contributions can be summarized as follows:
    %  \begin{itemize}
    %    \item We propose an unsupervised method (TS-LDDMM) to analyze inter-individual variability of shapes in a time series dataset. In particular, the method can handle \textit{irregularly sampled} time series with \textit{variable sizes}.
    %    \item We motivate our extension of LDDMM to time series by introducing a theoretical framework with a representation theorem for time series graph (\Cref{theorem:representation}) and kernels related to their structure (\Cref{lemma:choice_of_kernel_V}).
    %    \item We demonstrate the identifiability of the model by estimating the true generating parameter of synthetic data, and we highlight the sensitivity of our method with respect to its hyperparameters, also providing guidelines for tuning.
    %    We highlight the \textit{interpretability} of TS-LDDMM for studying the inter-individual variability in a clinical dataset.
    %     We illustrate the quantitative interest of the representation on classification tasks on real shape-based datasets.

    %  \end{itemize}
    Our contributions can be summarized as follows:
    \begin{itemize}
       \item We propose an unsupervised method (TS-LDDMM) to analyze the inter-individual variability of shapes in a time series dataset (\Cref{section:methodology}). In particular, the method can handle multivariate time series \textit{irregularly sampled} and with \textit{variable sizes}.
       \item We motivate our extension of LDDMM to time series by introducing a theoretical framework with a representation theorem for time series graph (\Cref{theorem:representation}) and kernels related to their structure (\Cref{lemma:choice_of_kernel_V}).
       \item We demonstrate the identifiability of the model by estimating the true generating parameter of synthetic data, and we highlight the sensitivity of our method concerning its hyperparameters (\Cref{appendix: settings_identifiability}), also providing guidelines for tuning (\Cref{appendix:kernel_TS_LDDMM}).
       \item We highlight the \textit{interpretability} of TS-LDDMM for studying the inter-individual variability in a clinical dataset (\Cref{section:experiments}).
        \item We illustrate the quantitative interest of such representation on classification tasks on real shape-based datasets with regular and irregular sampling (\Cref{appendix: robustness}-\ref{appendix: settings_classification}).
 \end{itemize}

    %To the best of our knowledge, these theoretical and practical works have never been done before.
      % On synthetic data, we show the sensibility of the results according to the kernel parameters and give guidelines to tune it.
      % In an unsupervised framework, we vectorize respitory patterns of a mouse dataset and analyze the variations depending on the drug taken by the mouse.
      % Finally, we compare to the state-of-the-art methods in classification an ... classifier taking as inputs the representation given by LDDMM.

  %    \paragraph*{Related papers}
  %    %https://arxiv.org/pdf/2106.00750.pdf look to the related papers
  %  In the shape analysis community, a famous paper \cite{srivastava2010shape} addresses the problem of representation of curves $C:\msi \to \Rset^d$ by using the Square-Root Velocity (SRV) representation.
  %      This structure is particularly relevant to compute geodesics and distances between curves when the space is quotiented by time reparametrization.
  %       However, in our case the time axis deserves a particular attention, since the curves are derived from time-series. 
  %       %While \cite{srivastava2010shape} tackles the computations of geodesics and distances on quotient curves space, our aim is to get a vector representation of each pattern. 
  %       Very recently, in \cite{heo2024logistic}, this representation was used to compute PCA and to perform classification on quotient curves space, we follow the same ideas, but we focus on time series and we use LDDMM instead of SRV.
 % optimization framework with a well chosen neural architecture in order to learn automatically the features, which has been very 
% To understand these complex data, we are looking to simpler representations toperform classical statiscal analysis or to perform efficient downstream tasks of machine learning, the time series representation res time series to a real valued vector representation. in order to apply classical statiscal or machine learning tools depending on the needs such as Principal Component Analysis (PCA), logistic regression or deep learning methods for classification or anomaly detection.

% \thi{
%   Our goal is to analyze the inter-individual variability within a time series dataset, an approach of prime interest in physiological contexts \cite{guscelli2019importance,wang2016research,bar2012studying,germain2023unsupervised}. More specifically, we aim to find an unsupervised features representation method that encodes individual time series specificities compared to a reference one.
% }

% \thi{
%   Studying shape differences between time series related to biological mechanisms is a common practice in physiology to characterize healthy and pathological functioning CITE. For instance, the shapes of heartbeats in electrocardiograms are discriminant for some cardiovascular pathologies CITE. Several approaches have been proposed for such comparison. Some employ shape-based similarity measures between time series \cite{asgari2023clustering,germain2023unsupervised,paparrizos2015k,ye2009time},  others embed time series as vectors of predefined features CITE, and, with the rise of deep neural networks, unsupervised learning representation of time series~\cite{trirat2024universal,meng2023unsupervised} has shown to be a valuable approach CITE notably with contrastive learning~\cite{franceschi2019unsupervised,tonekaboni2021unsupervised,meng2023unsupervised}. However, shape-based representation of time series within cohorts~\cite{niennattrakul2007inaccuracies,shirato2023identifying} remains an open problem in URL.
% }

% \thi{
%   Our work focuses explicitly on learning shape-based representation of time series. First, we propose not to see the shape of a time series through its curve $\{s_t:\eqsp t\in\msi\}$, but rather through its graph $\msg(s)=\{(t,s(t)):\eqsp t\in \msi\}$.
%    Then, building on the shape analysis literature \cite{beg2005computing,vaillant2004statistics}, we follow the Large Deformation Diffeomorphic Metric Mapping (LDDMM) framework \cite{beg2005computing,vaillant2004statistics} to analyze these graphs. The idea is to represent each element $\msg(s^j)$ of a dataset $(s^j)_{j\in[N]}$ as the transformation of a reference graph $\msg(\mathbf{s}_0)$ by a diffeomorphism $\phi_j$ (ie $\msg(s^j) \sim \phi_j . \msg(\mathbf{s}_0)$ ).
%     The diffeomorphism $\phi_j$ is learned by integrating an ordinary differential equation parameterized by a Reproducing Kernel Hilbert Space (RKHS).
%      The parameters $(\alpha_j)_{j\in[N]}$ encoding the diffemorphisms $(\phi_j)_{j\in[N]}$ yield the representation features of the graphs $(\msg(s^j))_{j\in[N]}$. Finally, these shape-encoding features can feed any statistical or machine-learning model.
%   }

% \thi{
%   However, a graph time series transformation by a general diffeomorphism is not always a graph time series, see e.g. \Cref{fig:diffeo}, thus a graph time series is more than a simple curve \cite{glaunes2008large}. Our contributions arise from this observation: we specify the class of diffeomorphisms to consider and show how to learn them. This change is fruitful in representing transformations of time series graphs as illustrated in \Cref{fig:transport}.
% }

% \thi{
%   Our contributions can be summarized as follows:
%     \begin{itemize}
%        \item We propose an unsupervised method (TS-LDDMM) to analyze the inter-individual variability of shapes in a time series dataset. In particular, the method can handle multivariate time series \textit{irregularly sampled} and with \textit{variable sizes}.
%        \item We motivate our extension of LDDMM to time series by introducing a theoretical framework with a representation theorem for time series graph (\Cref{theorem:representation}) and kernels related to their structure (\Cref{lemma:choice_of_kernel_V}).
%        \item We demonstrate the identifiability of the model by estimating the true generating parameter of synthetic data, and we highlight the sensitivity of our method concerning its hyperparameters, also providing guidelines for tuning.
%        \item We highlight the \textit{interpretability} of TS-LDDMM for studying the inter-individual variability in a clinical dataset.
%         \item We illustrate the quantitative interest of such representation on classification tasks on real shape-based datasets.
%  \end{itemize}
% }

\begin{figure}[t]
  \centering
  \includegraphics[width=0.7\linewidth]{"./pictures/diffeo.jpeg"}

  \caption{A time series' graph $\msg=\{(t,s(t)): \eqsp t\in\msi\} $ can lose its structure after applying a general diffeomorphism $\phi.\msg$: a time value can be related to two values on the space axis.}
  \label{fig:diffeo}

\end{figure}

@@

  \vspace{-1ex}
\section{Notations}
We denote by integer ranges by $[k:l]=\{k,\ldots,l\}\subset \mathcal{P}(\Zset)$ and $ [l]=[1:l]$ with $k,l\in \Nset$,
by $\rmC^m(\msi,\mse)$ the set of $m$-times continously differentiable function defined on an open set $\msu$ to a normed vector space $\mse$,
 by $||u||_\infty=\sup_{x\in \msu} |u(x)| $ for any bounded function $u:\msu \to \mse$,
and by $\Nset_{>0}$ is the set of positive integers. 
%We are in a sub class of curve quotiented by temporal reparametrisation since a time series' graph can be understood as the equivalent class of t ->(t,s(t))

%Reproducing Kernel Hilbert Space (RKHS)
\vspace{-1ex}
\section{Background on LDDMM}
\vspace{-1ex}
\label{section:LDDMM}

@@

 \vspace{-1ex}
% LDDMM is designed to analyzed an existing dataset, while NF and CNF are made to generalize a dataset for data augmentation.
\paragraph{Representing diffeomorpshims as deformations.}
Such \textit{deformations} $\phi$ are constructed via differential flow equations, for any $x_0\in \Rset^{d'} $ and $\tau\in[0,1]$:
\begin{equation}
  \label{eq:LDDMM_dynamic}
    \frac{\dd X(\tau)}{\dd \tau}= v_\tau(X(\tau)), \quad X(0)=x_0\eqsp ,
    \phi^v_\tau(x_0)=X(\tau), \quad \phi^v=\phi^v_1  \eqsp ,
\end{equation}
where the velocity field is $v:\tau\in [0,1]\mapsto v_\tau\in \msv $
and $\msv$ is a Hilbert space of continuously differentiable function
on $\Rset^{d'}$.  If
$||\dd u ||_{\infty}+|| u ||_{\infty}\leq ||u ||_\msv $ for any
$u\in \msv$ and
$v\in \rml^2([0,1],\msv)=\{v\in \rmC^0([0,1],\msv): \int_0^1 ||
v_\tau||^2_\msv \dd \tau<\infty \} $, by \citep[Theorem 5]{glaunes2005transport}
$\phi^v$ exists and belongs to $\mcd(\Rset^{d'})$, where we denote by $\mcd(\mso) $ the set of diffeomorpshim defined on an open set $\mso$ to $\mso$.
 Therefore, for any choice of $v$, $\phi^v$ defines a valid deformation. 
This offers a general recipe to construct diffeomorphism given a functional space $\msv$.

With this in mind, the velocity field $v$ fitting the data can be
estimated by minimizing 
$v \in \rml^2([0,1],\msv) \mapsto \mathscr{L}(\phi^{v}.\mathbf{x},\mathbf{y})$, where $\mathscr{L}$ is an appropriate loss function.
 However, two computational challenges arise.
  First, this optimization problem is ill-posed, and a penalty term is needed to obtain a unique solution.
   In addition, a parametric family $\msv_{\Theta} \subset \rml^2([0,1],\msv)$, parameterized by $\Theta$, is sought to efficiently solve this minimization problem. 
\paragraph*{From deformations to geodesics.}
%\paragraph*{Regularizing the problem to derive uniqueness.}
It has been proposed in \cite{miller2006geodesic} to interpret $\msv$ as a tangent space relative to the group of diffeomorphisms $\msh=\{ \phi^v:\eqsp v\in \rml^2([0,1],\msv)\}$.
Following this geometric point of view, geodesics can be constructed on $\msh$ by using the following squared norm 
 \begin{equation}
  \label{eq:geodesics_original}
    \mathscr{R}^2: g\in \msh\mapsto \inf_{ v\in \rml^2([0,1],\msv):\eqsp g=\phi^v} \int_0^1 || v_\tau||_\msv^2\dd \tau
 \end{equation}
By deriving differential constraints related to the minimum of \eqref{eq:geodesics_original} and using Cauchy-Lipschitz conditions, geodesics can be defined only by giving the starting point and the initial velocity $v_0\in \msv$ \cite{miller2006geodesic}, as straight lines in Euclidean space.
Denoting by $\tau \mapsto \rho_{v_0}(\tau)\in\msh$ the geodesic starting from the identity with inital velocity $v_0\in \msv$, the exponential map is defined as $\varphi^{\{v_0\}}\triangleq \rho_{v_0}(1)$. Using $\varphi^{\{v_0\}}$ instead of $\phi^v$, the previous matching problem becomes a \textit{geodesic shooting problem}:
 \begin{equation}
  \label{eq:geodesics_shooting}
  \inf_{v_0 \in \msv} \mathscr{L}(\varphi^{\{v_0\}}.\mathbf{x},\mathbf{y}).
 \end{equation}
 Using $\varphi^{\{v_0\}}$ instead of $\phi^v$ for any $v\in \rml^2([0,1],\msv)$ regularizes the problem and induces a sparse representation for the learning diffeomorphisms.
 Moreover, by setting $\msv$ as an RKHS, the geodesic shooting problem has a unique solution and becomes tractable, as described in the next section.

% More precisely, on the group of diffeomorphisms $\msh=\{ \phi^v:\eqsp v\in \rml^2([0,1],\msv)\}$,
%  the following squared norm can be defined
%  \begin{equation}
%   \label{eq:geodesics_original}
%     \mathscr{R}^2: g\in \msh\mapsto \inf_{ v\in \rml^2([0,1],\msv):\eqsp g=\phi^v} \int_0^1 || v_\tau||_\msv^2\dd \tau
%  \end{equation}
%      as the minimal "energy" needed to perform the deformation $g$.
% By \citep[Theorem 6]{glaunes2005transport}, there exists $v^*\in \rml^2([0,1],\msv)$ such that the previous infimum is a minimum in $v^*$ such that $(\phi^{v^*}_\tau)_{\tau\in[0,1]}$ can be understood as the geodesic between the identity function and $g$.
% However, given a diffeomorphism $g$, computing $\mathscr{R}(g)$ is intractable in most cases. 
% To circumvent this issue, another characterization of geodesics can be considered.
%  As in Riemannian geometry, 
%  instead of defining geodesics from their starting and end points, it is possible to define them from their starting point (here, the identity function) and an initial velocity $v_0\in\msv$.
%  \sam{To CHANGE}
%  %In other words, an Exponential map on deformations is wanted:
% given $v_0\in\msv$, it has been suggested to generate diffeomorphisms as $\varphi^{\{v_0\}}=\phi^v$ with
% %  \begin{equation}
% %   \label{eq:geodesics_shooting}
% %   v=\underset{w\in \rml^2([0,1],\msv):\eqsp w_0=v_0}{\argmin} \int_0^1 || w_\tau||_\msv^2\dd \tau\eqsp ,
% %  \end{equation}
% % since with this definition it holds $\mathscr{R}^2(\phi^v)= \int_0^1 || v_\tau||_\msv^2\dd \tau $.
%  By setting $\msv$ as an RKHS, the geodesic shooting problem \eqref{eq:geodesics_shooting} has a unique solution and becomes tractable, as described in the next section.
% Moreover, we have $\mathscr{R}(\varphi^{\{v_0\}})=||v_0||_\msv$ such that we can work with $v_0\in \msv $ instead of $v\in \rml^2([0,1],\msv)$.
%  Therefore, compared to CNF, only $v_0$ is parametrized by using the geodesic structure and not each $v_\tau$ for any $\tau\in[0,1]$.
%  Moreover, we have $\mathscr{R}(\phi^{\{v_0\}})=||v_0||_\msv$ such that we can work with $v_0\in \msv $ instead of $v\in L^2([0,1],\msv)$.
%  %Therefore, compared to CNF, only $v_0$ is parametrized by using the geodesic structure and not each $v_\tau$ for any $\tau\in[0,1]$.
%  This is presented with more details in the next paragraph.

@@

   %Before to specify the choice of the RKHS kernel $K$ and the loss on sets $\mathscr{L}$ in \Cref{section:time_series_specificity}, we first discuss the connection of LDDMM with normalizing flows. % to highlight the interest of the LDDMM approach.

%Conversely to NF or CNF, by \eqref{eq:integration}, different momentums parameters raise different final deformations $\phi^{\{v_0\}}$, this injectivity properties offers interpretability of the representation and gives sens to PCA or LDA post-processing as exposed in \Cref{section:experiments}.

   %If a diffeomorpshim $\Phi$ is given and we want to represent it as a deformation $\phi^v$ with $v$ of minimal ernergy, then
  %$(\alpha_{k,0})_k$ is the only parameters to learn since $(x_{k,0})_k$ is fixed.

%   \begin{algorithm}[tb]
%     \caption{Geodesic shooting with LDDMM}
%     \label{alg:example}
%  \begin{algorithmic}
%     \STATE {\bfseries Input:} a RKHS's kernel $K$, a loss on sets $L$, $\mathbf{x}=(x_{i,0})_{i\in[T_1]}$, $\mathbf{y}=(y_i)_{i\in[T_2]}$ two sets of $\Rset^{d'}$, initial momentums 
%     $(\alpha_{k,0})_{k\in [T_1]} \in (\Rset^{d'})^{T_1}$ and a number of gradient iterations $n_{\text{iter}}$.

%     \STATE $f: (\alpha_k)_k\mapsto L\left(\phi^{\{v_0\}}.\mathbf{x},\mathbf{y}\right)+\lambda||v_0||_\msv^2  $ with $v_0$ and $\phi^{\{v_0\}}$
%      respectively defined by \eqref{eq:def_v0} and \eqref{eq:integration} using control points $\mathbf{x}$ and initial momentums $(\alpha_k)_k $. 
%      \sam{Peut être à définir dans un autre algo ?}
%     \STATE $\alpha_{k,0}^*\gets \alpha_{k,0}, \eqsp k\in[T_1]$
%     \FOR{$i=1$ {\bfseries to} $n_{\text{iter}}$}
%     \STATE $\alpha_{k,0}^*\gets \text{AdamStep}(f,\alpha_{k,0}^*$) , where $\text{AdamStep}$ is given in appendix (mettre ref)
%     \ENDFOR
%     \STATE {\bfseries Return:} $(\alpha_{k,0}^*)_{k\in [T_1]}$

%  \end{algorithmic}
%  \end{algorithm}
\vspace{-1ex}
\paragraph{Relation to Continuous Normalizing Flows.}

One particular popular choice to address the problem of learning a diffeomorphism or a velocity field is Normalizing Flows \cite{rezende2015variational,kobyzev2020normalizing} (NF) or their continuous counterpart \cite{chen2018neural,grathwohl2019scalable,salman2018deep} (CNF).
However, we do not rely on this class of learning algorithms for several reasons. Indeed, existing and simple normalizing flows are not suitable for the type of data that we are interested in this paper \cite{feng2023multi,deng2020modeling}. 
  In addition,  they are primarily designed to have tractable Jacobian functions, while we do not require such property in our applications. 
Finally, the use of a differential flow solution of an ODE
\eqref{eq:LDDMM_dynamic} trick is also at the basis of CNF, which
then consists of learning a velocity field to address in fitting the data through a loss aiming to address the problem
at hand. Nevertheless, the main difference between CNF and LDDMM lies in the
parametrization of the velocity field. LDDMM uses kernels to
derive closed form formula and enhance interpretability while
NF and CNF take advantage of deep neural networks to scale with
large dataset in high dimensions.
\vspace{-1ex}
\section{Methodology}
\label{section:methodology}
%\paragraph{Assumptions}
%Denoting by $C^0(\msi,\msj) $ the space of continous functions between two sets $\msi,\msj$,
\vspace{-1ex}
%\sam{Lien avec functional data analysis ? Insister sur le fait que concaténer temps et espace c'est moisi}
We consider in this paper observations which consist in a population of $N$ multivariate time series, for any $j\in[N]$, $s^j \in \rmC^1(\msi_j,\Rset^{d})$. 
However, we can only access a $n_j$-samples $\tsig^j=(\tsig_i^j=s^j(t^j_i))_{i\in[n_j]}$ collected at timestamps $(t^j_i)_{i\in[n_j]}$ for any $j \in [N]$.
 Note that \textbf{the number of samples $n_j$ is not necessarily the same across individuals} and the timestamps can be \textbf{irregularly sampled}.
 We assume the time series population is globally homogeneous regarding their "shapes" even if inter-individual variability exists.

>>>>

\vspace{-1ex}
\section{Introduction}
\vspace{-1ex}

@@

Our work explicitly focuses on learning shape-based representation of time series. First, we propose to view the shape of a time series not merely as its curve $\{s_t:\eqsp t\in\msi\}$, but as its graph $\msg(s)=\{(t,s(t)):\eqsp t\in \msi\}$. Then, building on the shape analysis literature \cite{beg2005computing,vaillant2004statistics}, we adopt the Large Deformation Diffeomorphic Metric Mapping (LDDMM) framework \cite{beg2005computing,vaillant2004statistics} to analyze these graphs. The core idea is to represent each element $\msg(s^j)$ of a dataset $(s^j)_{j\in[N]}$ as the transformation of a reference graph $\msg(\mathbf{s}_0)$ by a diffeomorphism $\phi_j$, i.e. $\msg(s^j) \sim \phi_j . \msg(\mathbf{s}_0)$. The diffeomorphism $\phi_j$ is learned by integrating an ordinary differential equation parameterized by a Reproducing Kernel Hilbert Space (RKHS). The parameters $(\alpha_j)_{j\in[N]}$ encoding the diffemorphisms $(\phi_j)_{j\in[N]}$ yield the representation features of the graphs $(\msg(s^j))_{j\in[N]}$. Finally, these shape-encoding features can be used as inputs to any statistical or machine-learning model.

However, a graph time series transformation by a general diffeomorphism is not always a graph time series, see e.g. \Cref{fig:diffeo}, thus a graph time series is more than a simple curve \cite{glaunes2008large}. Our contributions arise from this observation: we specify the class of diffeomorphisms to consider and show how to learn them. This change is fruitful in representing transformations of time series graphs as illustrated in \Cref{fig:transport}.

Our contributions can be summarized as follows:
\begin{itemize}
  \item We propose an unsupervised method (TS-LDDMM) to analyze the inter-individual variability of shapes in a time series dataset (\Cref{section:methodology}). In particular, the method can handle multivariate time series \textit{irregularly sampled} and with \textit{variable sizes}.

  \item We motivate our extension of LDDMM to time series by introducing a theoretical framework with a representation theorem for time series graph (\Cref{theorem:representation}) and kernels related to their structure (\Cref{lemma:choice_of_kernel_V}).

  \item We demonstrate the identifiability of the model by estimating the true generating parameter of synthetic data, and we highlight the sensitivity of our method concerning its hyperparameters (\Cref{appendix: settings_identifiability}), also providing guidelines for tuning (\Cref{appendix:kernel_TS_LDDMM}).

  \item We highlight the \textit{interpretability} of TS-LDDMM for studying the inter-individual variability in a clinical dataset (\Cref{section:experiments}).

  \item We illustrate the quantitative interest of such representation on classification tasks on real shape-based datasets with regular and irregular sampling (\Cref{appendix: robustness,appendix: shape_classification}).
\end{itemize}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.7\linewidth]{"./pictures/diffeo.jpeg"}
  \caption{A time series' graph $\msg=\{(t,s(t)): \eqsp t\in\msi\} $ can lose its structure after applying a general diffeomorphism $\phi.\msg$: a time value can be related to two values on the space axis.}
  \label{fig:diffeo}
\end{figure}

@@

  \vspace{-1ex}
\section{Notations}
We denote by integer ranges by $[k:l]=\{k,\ldots,l\}\subset \mathcal{P}(\Zset)$ and $ [l]=[1:l]$ with $k,l\in \Nset$, by $\rmC^m(\msi,\mse)$ the set of $m$-times continously differentiable function defined on an open set $\msu$ to a normed vector space $\mse$, by $||u||_\infty=\sup_{x\in \msu} |u(x)| $ for any bounded function $u:\msu \to \mse$, and by $\Nset_{>0}$ is the set of positive integers.

\vspace{-1ex}
\section{Background on LDDMM}
\vspace{-1ex}
\label{section:LDDMM}

@@

\vspace{-1ex}
\paragraph{Representing diffeomorpshims as deformations.}
Such \textit{deformations} $\phi$ are constructed via differential flow equations, for any $x_0\in \Rset^{d'} $ and $\tau\in[0,1]$:
\begin{equation}
  \label{eq:LDDMM_dynamic}
    \frac{\dd X(\tau)}{\dd \tau}= v_\tau(X(\tau)), \quad X(0)=x_0\eqsp ,
    \phi^v_\tau(x_0)=X(\tau), \quad \phi^v=\phi^v_1  \eqsp ,
\end{equation}
where the velocity field is $v:\tau\in [0,1]\mapsto v_\tau\in \msv $ and $\msv$ is a Hilbert space of continuously differentiable function on $\Rset^{d'}$. If $||\dd u ||_{\infty}+|| u ||_{\infty}\leq ||u ||_\msv $ for any $u\in \msv$ and $v\in \rml^2([0,1],\msv)=\{v\in \rmC^0([0,1],\msv): \int_0^1 ||v_\tau||^2_\msv \dd \tau<\infty \} $, by \citep[Theorem 5]{glaunes2005transport} $\phi^v$ exists and belongs to $\mcd(\Rset^{d'})$, where we denote by $\mcd(\mso) $ the set of diffeomorpshim defined on an open set $\mso$ to $\mso$. Therefore, for any choice of $v$, $\phi^v$ defines a valid deformation. This offers a general recipe to construct diffeomorphism given a functional space $\msv$.

With this in mind, the velocity field $v$ fitting the data can be estimated by minimizing $v \in \rml^2([0,1],\msv) \mapsto \mathscr{L}(\phi^{v}.\mathbf{x},\mathbf{y})$, where $\mathscr{L}$ is an appropriate loss function. However, two computational challenges arise. First, this optimization problem is ill-posed, and a penalty term is needed to obtain a unique solution. In addition, a parametric family $\msv_{\Theta} \subset \rml^2([0,1],\msv)$, parameterized by $\Theta$, is sought to efficiently solve this minimization problem.

\vspace{-1ex}
\paragraph{From deformations to geodesics.}
It has been proposed in \cite{miller2006geodesic} to interpret $\msv$ as a tangent space relative to the group of diffeomorphisms $\msh=\{ \phi^v:\eqsp v\in \rml^2([0,1],\msv)\}$. Following this geometric point of view, geodesics can be constructed on $\msh$ by using the following squared norm 
\begin{equation}
  \label{eq:geodesics_original}
  \mathscr{R}^2: g\in \msh\mapsto \inf_{ v\in \rml^2([0,1],\msv):\eqsp g=\phi^v} \int_0^1 || v_\tau||_\msv^2\dd \tau
\end{equation}
By deriving differential constraints related to the minimum of \eqref{eq:geodesics_original} and using Cauchy-Lipschitz conditions, geodesics can be defined only by giving the starting point and the initial velocity $v_0\in \msv$ \cite{miller2006geodesic}, as straight lines in Euclidean space. Denoting by $\tau \mapsto \rho_{v_0}(\tau)\in\msh$ the geodesic starting from the identity with inital velocity $v_0\in \msv$, the exponential map is defined as $\varphi^{\{v_0\}}\triangleq \rho_{v_0}(1)$. Using $\varphi^{\{v_0\}}$ instead of $\phi^v$, the previous matching problem becomes a \textit{geodesic shooting problem}:
\begin{equation}
  \label{eq:geodesics_shooting}
  \inf_{v_0 \in \msv} \mathscr{L}(\varphi^{\{v_0\}}.\mathbf{x},\mathbf{y}).
\end{equation}
Using $\varphi^{\{v_0\}}$ instead of $\phi^v$ for any $v\in \rml^2([0,1],\msv)$ regularizes the problem and induces a sparse representation for the learning diffeomorphisms. Moreover, by setting $\msv$ as an RKHS, the geodesic shooting problem has a unique solution and becomes tractable, as described in the next section.

@@

\vspace{-1ex}
\paragraph{Relation to Continuous Normalizing Flows.}
One particular popular choice to address the problem of learning a diffeomorphism or a velocity field is Normalizing Flows \cite{rezende2015variational,kobyzev2020normalizing} (NF) or their continuous counterpart \cite{chen2018neural,grathwohl2019scalable,salman2018deep} (CNF). However, we do not rely on this class of learning algorithms for several reasons. Indeed, existing and simple normalizing flows are not suitable for the type of data that we are interested in this paper \cite{feng2023multi,deng2020modeling}. In addition,  they are primarily designed to have tractable Jacobian functions, while we do not require such property in our applications. Finally, the use of a differential flow solution of an ODE \eqref{eq:LDDMM_dynamic} trick is also at the basis of CNF, which then consists of learning a velocity field to address in fitting the data through a loss aiming to address the problem at hand. Nevertheless, the main difference between CNF and LDDMM lies in the parametrization of the velocity field. LDDMM uses kernels to derive closed form formula and enhance interpretability while NF and CNF take advantage of deep neural networks to scale with large dataset in high dimensions.

\vspace{-1ex}
\section{Methodology}
\label{section:methodology}
\vspace{-1ex}

We consider in this paper observations which consist in a population of $N$ multivariate time series, for any $j\in[N]$, $s^j \in \rmC^1(\msi_j,\Rset^{d})$. However, we can only access a $n_j$-samples $\tsig^j=(\tsig_i^j=s^j(t^j_i))_{i\in[n_j]}$ collected at timestamps $(t^j_i)_{i\in[n_j]}$ for any $j \in [N]$. Note that \textbf{the number of samples $n_j$ is not necessarily the same across individuals} and the timestamps can be \textbf{irregularly sampled}. We assume the time series population is globally homogeneous regarding their "shapes" even if inter-individual variability exists. Intuitively speaking, the "shape" of a time series $s:\msi\to \Rset^d$ is encoded in its graphs $\msg(s)$ defined as the set $\{(t,s(t)):\eqsp t\in\msi \} $ and not only in its values $s(\msi)=\{s(t):\eqsp t\in\msi \} $ since the time axis is crucial. As a motivating use-case, $s^j$ can be the time series of a heartbeat extracted from an individual's electrocardiogram (ECG), see \Cref{fig:transport}. The homogeneity in a resulting dataset comes from the fact that humans have similar shapes of heartbeat \cite{ye2012heartbeat,madona2021pqrst}.

\vspace{-1ex}
\paragraph*{The deformation problem.}
In this paper, we aim to study the inter-individual variability in the dataset by finding a relevant representation of each time series. Inspired from the framework of shape analysis \cite{vaillant2004statistics}, addressing similar problems in morphology, we suggest to represent each time series' graph $\msg(s^j)$ as the transformation of a reference graph $\msg(\mathbf{s}_0)$, related to a time series $\mathbf{s}_0:\msi \to\Rset^d$, by a diffeomorphism $\phi_j$ on $\Rset^{d+1}$, for any $j\in[N]$,
\begin{equation}
 \label{eq:transformation}
 \phi_j.\msg(\mathbf{s}_0)=\{\phi_j\left(t,\mathbf{s}_0(t)\right), \eqsp t\in \msi \} \eqsp.
\end{equation}
$\bfs_0$ will be understood as the typical representative shape common to the collection of time series $(s^j)_{j\in[N]}$. As $\bfs_0$ is supposed to be fixed, then the representation of the time series $(s^j)_{j\in[N]}$ boils down to the one of the transformation $(\phi_j)_{j\in[N]}$. We aim to learn $\msg(\bfs_0)$ and $(\phi_j)_{j\in[N]} $.
